Safe Redis Queries

### 2016-06-12 aplas16/sections/IndexedMonads.tex
Half of Section 3.3.

Stateful computations are often reasoned in a Hoare-logic style: each command
is labelled by a \emph{precondition} and a \emph{postcondition}. If the former
is satisfied before the command is executed, the latter is guaranteed to hold
afterwards.

In Haskell, stateful computations are represented by monads. In order to
reason about their behaviors within the type system, we wish to label a state
monad with its pre and postcondition. An \emph{indexed
monad}~\cite{indexedmonad} (also called \emph{monadish} or
\emph{parameterised monad}) is a monad that, in addition to the type of value
it computes, takes two more type arguments representing an initial state and
a final state, to be interpreted like a Hoare triple~\cite{kleisli}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The intention is that a computation of type \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is a stateful computation
such that if it starts execution in a state satisfying \ensuremath{\Varid{p}} and terminates, it
yields a value of type \ensuremath{\Varid{a}}, and the new state satisfies \ensuremath{\Varid{q}}.
The operator \ensuremath{\Varid{unit}} lifts a pure computation to a stateful computation that
does not alter the state. In \ensuremath{\Varid{x}\mathbin{`\Varid{bind}`}\Varid{f}}, a computation \ensuremath{\Varid{x}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} must
be chained before \ensuremath{\Varid{f}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}}, which expects a value of type \ensuremath{\Varid{a}} and
a state satisfying \ensuremath{\Varid{q}} and, if terminates, ends in a state satisfying \ensuremath{\Varid{r}}.
The result is a monad \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}} --- a computation that, if executed in a state
satisfying \ensuremath{\Varid{p}} and terminates, yields a value \ensuremath{\Varid{b}} and a state satisfying \ensuremath{\Varid{r}}.
Indexed monads have been used ~\cite{typefun,staticresources} ... \todo{for what? Some discriptions here to properly cite them.}

We define a new indexed monad \ensuremath{\Conid{Popcorn}} which, at term level, merely wraps
\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the pre and
postconditions at type level:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to \ensuremath{\Varid{unPopcorn}} to erase the additional type information and get back an ordinary \Hedis{} program.

\paragraph{\text{PING}: A First Example}
In \Redis{}, \text{PING} does nothing but replies with
 \text{PONG} if the connection is alive. In Hedis,
 \ensuremath{\Varid{ping}} has type:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Now with \ensuremath{\Conid{Popcorn}}, we could make our own version of
\ensuremath{\Varid{ping}}\footnotemark

\footnotetext{\ensuremath{\Varid{ping}} from Hedis is qualified with
\ensuremath{\Conid{Hedis}} to prevent function name clashing in our code.}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Popcorn}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The dictionary \ensuremath{\Varid{xs}} in the type remains unaffected after the
 action, because \ensuremath{\Varid{ping}} does not affect any key-type
 bindings. To encode other commands that modifies key-type bindings, we need
 type-level functions to annotate those effects on the dictionary.

>>>>

Stateful computations are often reasoned using Hoare logic. A {\em Hoare triple}
$\{P\} S \{Q\}$ denotes such a proposition: if the statement $S$ is executed in
a state satisfying prediate $P$, when it terminates, the state must satisfy
predicate $Q$. Predicates $P$ and $Q$ are respectively is called the
\emph{precondition} and the \emph{postcondition} of the Hoare triple.

In Haskell, stateful computations are represented by monads. In order to
reason about their behaviors within the type system, we wish to label a state
monad with its pre and postcondition. An \emph{indexed monad}~%
\cite{indexedmonad} (also called \emph{monadish} or \emph{parameterised monad})
is a monad that, in addition to the type of value it computes, takes two more
type arguments representing an initial state and a final state, to be
interpreted like a Hoare triple~\cite{kleisli}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The intention is that a computation of type \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is a stateful computation
such that, if it starts execution in a state satisfying \ensuremath{\Varid{p}} and terminates, it
yields a value of type \ensuremath{\Varid{a}}, and the new state satisfies \ensuremath{\Varid{q}}. The operator \ensuremath{\Varid{unit}}
lifts a pure computation to a stateful computation that does not alter the
state. In \ensuremath{\Varid{x}\mathbin{`\Varid{bind}`}\Varid{f}}, a computation \ensuremath{\Varid{x}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is followed by
\ensuremath{\Varid{f}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}} --- the postcondition of \ensuremath{\Varid{x}} matches the precondition of
the computation returned by \ensuremath{\Varid{f}}. The result is a monad \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}}.
Indexed monads have been used ~\cite{typefun,staticresources} ... \todo{for what? Some discriptions here to properly cite them.}

We define a new indexed monad \ensuremath{\Conid{Popcorn}} which, at term level, merely wraps
\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the
pre/postconditions at type level:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
At term level, the \ensuremath{\Varid{unit}} and \ensuremath{\Varid{bind}} methods are not interesting: they merely
make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of \ensuremath{\Conid{Redis}}, and extracts and re-apply the constructor \ensuremath{\Conid{Popcorn}} when necessary. With \ensuremath{\Conid{Popcorn}} being a \ensuremath{\mathbf{newtype}}, they
can be optimized away in runtime. The interesting bits happen in compile type,
on the added type information.

The properties of the state we care about are the set of currently allocated
keys and their associated types. We will present, in Section~\ref{sec:type-level-dict}, techniques that allow us to specify
properties such as ``the keys in the database are \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}},
respectively associated to values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and \ensuremath{\Conid{Bool}}.''
For now, however, let us look at the simplest \Redis{} command.

The command \text{PING} in \Redis{} does nothing but replies a message
\text{PONG} if the connection is alive. In \Hedis{}, \ensuremath{\Varid{ping}} has type
\ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status})}. The \Popcorn{} version of \ensuremath{\Varid{ping}} simply
applys an additional constructor (functions from \Hedis{} are qualified with
\ensuremath{\Conid{Hedis}} to prevent name clashing):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Popcorn}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Since \ensuremath{\Varid{ping}} does not alter the database, the postcondition and precondition
are the same. Commands that are more interesting will be introduced after
we present our type-level encoding of constraints on states.
